---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
header-includes:
    - \usepackage{setspace}\doublespacing
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\tableofcontents
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

The political history of the United States has been characterized by a general, if unlinear, trend toward universal sufferage [see, for instance, @Keyssar2009]. At the time of the nation's founding, access to the ballot box was restricted to landed White men; over the following two centuries, the franchise was greatly expanded. Today, voting rights are considered foundational aspects of full citizenship in a liberal democracy. And yet, despite the United State's purported march toward ever-more-inclusive systems of democracy, one large group of American citizens is formally barred from voting. In most of the United States, citizens convicted of felonies are at least temporarily prohibited from casting ballots in elections [@bcj_laws]. Although some states such as Florida and Louisiana have gradually moved to dismantle their systems of felony disenfranchisement, an estimated `r 6.1 - 1.4` million American citizens remain disenfranchised today [@sentencing_2016]. These disenfranchising policies provide a stark reminder that our democracy is not as inclusive as we might like to tell ourselves. 

The disenfranchisement of citizens convicted of felony offenses is particularly troubling given the racialized and place-based patterns of policing and incarceration in the United States. As Michelle Alexander [-@Alexander2012] and others have shown, mass incarceration in the post-Civil Rights era has been a tool used by the state to exert control over minority --- and particularly Black --- Americans. This is clear in the case of states such as New York: according to data from the New York State Department of Corrections and Community Supervision, `r round(readRDS("./temp/share_black_incarcerated.rds"), digits = 3) * 100` percent of individuals who were incarcerated in December of 2018 were Black, although the Census Bureau estimates that just 14.6 percent of the citizen voting age population in the state is Black. This disparity is likely due *not* to any inherent differential propensity to commit crimes among different racial groups, but rather to systems of policing and concentrated poverty. As @Gelman2007 shows, for instance, New York's "stop-and-frisk" policy targeted Black and Latino New Yorkers at rates far higher than Whites, even after controlling for neighborhood variability and race-specific criminal propensity.

Due to economic and racial segregation, these effects are highly spatially concentrated. Data available from New York City shows that in 2017, 10 of the New York Police Department's 77 precincts were responsible for more than a quarter of all arrests for felony charges. Many scholars have detailed the impact of living in areas with high levels of police activity. Residents of such neighborhoods suffer from worse physical health [@Sewell2016] and are more likely to suffer from anxiety and exhibit symptoms of trauma [@Geller2014]. The labor markets and social networks in neighborhoods with high levels of policing and incarceration are disrupted [@Clear2008]. Concentrated policing has also credited with having a "chilling effect" on neighborhoods' willingness to reach out for help to local governments [@Lerman2013].

Felony disenfranchisement policies are part of an interlocking criminal justice system that disproportionately impacts Black Americans living in certain communities. The effects of disenfranchisement are concentrated in neighborhoods that already suffer from myriad disadvantanges stemming from targeted policing and economic marginalization. The neighborhood-specific implications of felony disenfranchisement, however, largely remain unstudied. A number of studies have explored the effect of imprisonment and disenfranchisement on later political participation [@White2019; @Burch2011; @Gerber2014]. Others have looked at the spillover effects of disenfranchisement on eligible Black voters at the state level [@Bowers2009; @King2016]. With the exception of @Burch2013, however, little attention has been paid to the impact of felony disenfranchisement on political participation at the neighborhood level.

This paper seeks to extend the extant literature on felony disenfranchisement by considering the implications of disenfranchisement on local political participation. I begin by looking at participation rates in New York City neighborhoods where individuals with a demonstrated history of participation have been disenfranchised. This serves as a refinement on past work, which considered only which individuals are formally disenfranchised, regardless of whether they had cast a ballot in the past. The disadvantaged state of neighborhoods with high levels of incarceration has been firmly established; if felony disenfranchisement results in lower turnout among elgibile voters in these communities, the literature has likely *understated* the degree to which they are impacted by spatially concentrated patterns of criminal justice.

Undoing the negative effects of felony disenfranchisement on individual and neighborhood representation requires crafting policy aimed at increasing turnout among the formerly disenfranchised. New York State has taken a step in that direction: in April of 2018, Governor Andrew Cuomo signed Executive Order 181, which came close to ending disenfranchisement for New Yorkers on parole. After examining the neighborhood effects of felony disenfranchisement, I test whether having one's rights restored while still on parole increased turnout among formerly disenfranchised individuals in the 2018 midterm elections.

# Background of Felony Disenfranchisement in the United States {-}

In all but two states (Maine and Vermont), felony disenfranchisement laws ensure that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor [@bcj_laws]. This variation in laws flows directly from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights "for participation in rebellion, or other crime." The definition of "other crime," left so vague in the Constitution, is now generally used by states to disenfranchise citizens for any felony offense. The Supreme Court, in cases such as *Richardson v. Ramirez* (1974), has upheld states' right to do just that. Collectively, these laws disenfranchise as many as `r 6.1 - 1.4` million American citizens. Of these, the majority are no longer incarcerated, but are living and working in their communities [@sentencing_2016].^[The figures reported in @sentencing_2016 have been adjusted to reflect the impact of Amendment 4 in Florida.]

In any discussion of felony disenfranchisement in the United States, it is imperative to acknowledge the central role played by race and white supremacy. As Traci Burch has explained, "If policies restricting the voting rights of offenders disparately affect one racial group or party, it is because such policies were *intended* to" [@Burch2010, p. 4; emphasis in the original]. Previous research has established that the presence of nonwhite potential voters is associated with the implementation of felony disenfranchisement policies and that these policies were often adopted during Jim Crow to reduce the political power of Black Americans [@Behrens2003]. In Florida, for instance, felony disenfranchisement was added to the state constitution in 1868. Afterwards, a lawmaker boasted that the amendment had been added to prevent the state from being "niggerized" [@florida_1868, p. 374].

The racially unbalanced effects of felony disenfranchisement laws were not confined to the 19th century. Although the Voting Rights Act of 1965 did much to improve access to the ballot box for minorities, it did nothing to undermine the explicitly racialized system of disenfranchisement. Indeed, as the United States has increased the reach of the carceral system in the post-Civil Rights era, the implications of felony disenfranchisement have only grown. As of 2016, more than 10 percent of Black Americans were disenfranchised in 9 states. In Kentucky, the state with the highest level of disenfranchised Black residents, more than one in four Black adults are barred from casting a ballot. Although Black adults made up just `r percent(29932674 / 247219588, accuracy = 0.1)` of the voting age population in 2016, they accounted for `r percent(2228118 / 6106327, accuracy = 0.1)` of the disenfranchised population [@sentencing_2016].

Recent data from Florida make the victims of felony disenfranchisement laws especially clear. Prior to January, 2019, Floridians convicted of felony offenses were permanently disenfranchised unless they were individually pardoned by a clemency board. In November, 2018, voters passed a ballot initiative amending the state constitution to end permanent disenfranchisement. In the first three months after permanent disenfranchisement was ended, 44 percent of formerly incarcerated Floridians who registered to vote were Black --- compared with just 13 percent of the statewide electorate [@Morris2019].

# Academic Literature and Felony Disenfranchisement {-}

In the aftermath of the 2000 presidential election, academic interest in the political implications of felony disenfranchisement was stirred thanks to a paper from @Uggen2002. George W. Bush's margin of victory in Florida in 2000 was famously just 537 votes. In their 2002 paper, Uggen and Manza estimate the likely partisan composition of the disenfranchised population with felony convictions in their past. They estimate that, if this group had been allowed to vote, they would have supported Al Gore by a wide margin. Their enfranchisement, Uggen and Manza argued, would have tipped the presidential contest and resulted in the election of Al Gore. They based their estimates on the voting patterns of eligible individuals who were demographically similar to the disenfranchised population. Though much of the research conducted since their 2002 study has pushed back against some of their key assumptions (namely, that formerly incarcerated individuals turn out at the same rate as those who have not been incarcerated), Uggen and Manza convincingly demonstrated that felony disenfranchisement can have material political consequences. In the years after the Uggen and Manza published their paper, scholars sought to investigate the relationship between felony disenfranchisement and Black and youth turnout [@Miles2004; @Hjalmarsson2010]. Some of this research compared states and regions with differing disenfranchisement regimes to estimate these effects [@Miles2004; @Ochs2006]. Others have used survey data or interviews to construct their estimates [@Uggen2004; @Drucker2005].

In a series of papers between 2009 and 2011, researchers developed methods for directly estimating the turnout of formerly disenfranchised individuals. @Haselswerdt2009 matched release data and voter registration data from Erie County, NY, to estimate turnout among a small group of formerly incarcerated individuals. Traci Burch [-@Burch2010; -@Burch2011] expanded upon this matching methodology to estimate the voting patterns of formerly disenfranchised individuals in a range of states. She used release data from states' Departments of Corrections and their registered voter files to identify formerly incarcerated individuals who went on to register to vote. Using the registered voter files, she was also able to estimate the party affiliation of formerly incarcerated individuals (in states with party registration) and their turnout rates. Her methodology has been used to investigate other questions surrounding the voting patterns of formerly incarcerated individuals under different circumstances and to examine the impact of changes in disenfranchisement policy [@Meredith2013; @Meredith2015].

A number of papers have also explored the impact that felony disenfranchisement policies have on turnout among non-disenfranchised residents. @King2016, for instance, leverages state-level variation in disenfranchisement laws to estimate the impact that felony disenfranchisement has on turnout among Black Americans. They use data from the 2004 Current Population Survey Voting and Registration Supplement to determine individual-level turnout. They include estimates of the share of Black Americans who are disenfranchised in each state from @locked_out to explore the impact of these policies on eligible voters. Ultimately, they argue that "African American disenfranchisement plays a unique role in predicting African American voter turnout" (p. 800). They conclude that disenfranchisement has spillover effects for Black voters: where more Black residents are disenfranchised, elibible Black voters are less likely to cast a ballot. These findings are in line with other research that has explored whether the effects of disenfranchisement extend beyond those whose voting rights are directly suspended [@Bowers2009; @Ochs2006; @Walker2014]. As @Bowers2009 sums up: "[I]t is not solely the direct vote of ex-felons that is denied through these laws. [Felony disenfranchisement] impacts the political power of communities that extends beyond felons' collateral penalty" (p. 724).

Although scholars have established that felony disenfranchisement decreases turnout among Black voters at the *state* level, relatively little research has been done on how felony disenfranchisement operates at the sub-state level. Though we know that Black voters are generally less likely to cast a ballot when they live in a state with strict disenfranchisement laws, less work has been done exploring the impact these laws might have at the local level. @Burch2013 is an exception to this. In this paper, Burch explores the depressive effect of disenfranchisement laws at the local level in North Carolina by examining census block-group level turnout and involvement with the criminal justice system. Using a standard ordinary least squares regression, she determined that "at high concentrations, imprisonment and community supervision have an unequivocally demobilizing effect of neighborhoods" (p. 185). This paper seeks to expand on her work by replicating her findings in New York City; by using a different estimation technique; and by using a different definiton of "lost voters."

# What This Paper Examines {-}

This paper begins by exploring the effect of felony disenfranchisement on neighborhood turnout in the New York City Mayoral election of 2017. Policing and incarceration patterns have historically targeted communities of color, wreaking havok on the social fabric of these neighborhoods [e.g. @Sewell2016; @Clear2008; @Lerman2013]. It is possible, however, that these effects run deeper than the direct focus on policing and incarceration has acknowledged. We know that felony disenfranchisement systematically removes voters from certain neighborhoods, but it is not clear whether enough voters are removed relative to the electorate to meaningfully distort neighborhood representation. To the extent that felony disenfranchisement has a depressive effect on *eligible* voters, it is possible that these policies are powerful enough to materially reduce the representatition of certain parts of the city.

After examining the effect of felony disenfranchisement on voter turnout at the neighborhood level, I consider the efficacy of a program intended to undo one piece of the felony disenfranchisement apparatus. In April, 2018, Governor Andrew Cuomo signed Executive Order 181 which restored voting rights to New Yorkers on parole.^[Prior to Executive Order 181, residents convicted of felonies in prison and on parole were barred from voting, while individuals sentenced to probation did not have their voting rights revoked. Today, only those in prison for felony offenses are disenfranchised.] Parole officers were required to provide registration forms the parolees under their supervision and inform them of their voting rights. In theory, such a policy could increase turnout and registration for parolees. To test whether the policy was successful, I build individual-level models to explore whether turnout rates in the 2018 midterm elections among individuals discharged from parole were impacted by the implementation of Executive Order 181.

The first part of this project seeks to refine how we think about felony disenfranchisement in the United States. Rather thank think about it purely through the prism of those directly impacted (residents who are actively disenfranchised) or through a prism of race (by examining the impact of disenfranchisement on Black turnout at the state level), I seek to introduce the importance of physical space into the conversation. This will allow us to both better understand the mechanisms through which the depressive effects operate and also to understand more deeply where these effects are concentrated. Of course, understanding the impact of felony disenfranchisement is not enough: we must also test whether our efforts to undo its legacy are effective. The second part of this project seeks to do just that.

## Framing for Turnout Effects {-}

As discussed above, it has been widely established that felony disenfranchisement reduces turnout even among eligible voters, particularly in the Black community [e.g. @King2016; @Bowers2009]. With the exception of @Burch2013, however, there has been little investigation into whether these depressive effects are narrowly concentrated in the neighborhoods home to disenfranchised individuals or are more widely dispersed. The degree to which these effects are concentrated in particular neighborhoods has important implications for our understanding of the problem of felony disenfranchisement. Although literature examining the degree to which geographic polarization effects voting patterns is scarce, there is some evidence to suggest that neighborhoods do vote similarly. @Kinsella2015, for instance, examines political clustering in the Greater Cincinnati Metropolitan Area from 1976 through 2008. Over this three-decade period, precinct-level presidential election results show trends toward increased polarization. Insofar as neighborhoods show distinct preferences for politicians, spatially concentrated depressive effects from felony disenfranchisement could influence election outcomes.

Even if neighborhoods to not have unique preferences for presidential candidates, lower turnout in impacted communities may reduce their allocation of public goods. At the Congressional level, for instance, representatives direct resources to areas within their districts that provide the greatest political benefit [@Martin2003]. Congressional representatives are also more responsive to the policy preferences of higher-turnout areas. "[H]igher citizen participation is rewarded," @Martin2013 concludes, "with enhanced policy responsiveness" (p. 59). @Griffin2005 finds similar effects in the United States Senate, reporting that "voter preferences predict the aggregate roll-call behavior of Senators while nonvoter preferences do not" (p. 1206). If the neighborhoods most impacted by felony disenfranchisement turn out at lower rates, they may find that their elected representatives are less likely to support their needs.

Although research on the impact of local turnout on city-wide policy is scarce, @Anzia2019 examines the impact of senior turnout on "senior-friendly" policy at the city level. Anzia does not find that senior turnout in general increases the likelihood of senior-friendly policies, but that elected officials *are* responsive to senior turnout when seniors "are a more cohesive, meaningful group." American cities are highly segregated by race and by class, but less so by age. As such, Anzia's study does not speak directly to the impact of *neighborhood* turnout rates on city policy. An extension of her paper, however, implies that what she finds might be probitive to the effect of neighborhood turnout rates. Insofar as neighborhoods vote as a bloc, their relative turnout rates may influence city policy. If felony disenfranchisement has a depressive effect on turnout, politicians are likely less responsive to the particular preferences of voters in these neighborhoods  --- some of the most neglected neighborhoods to begin with.

Despite over-incarceration in some neighborhoods, the number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with `r round(readRDS("./temp/active_voters_0319.rds")/1000000, digits = 1)` million actively registered voters. Despite the low share of residents who are directly disenfranchised, there is reason to believe the policy impacts more individuals than just those imprisoned. As discussed above, previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher, an effect often referred to as *de facto* disenfranchisement.

This *de facto* disenfranchisement is likely to be concentrated within the neighborhoods home to formally disenfranchised residents. Voting is a social act, and social networks play an important role in predicting political participation [e.g. @Foladare1968; @Huckfeldt1979; @Kenny1992; @Mutz2002]. Literature from urban sociology has established that social networks are largely spatially bounded, and that local social ties are more important in lower-income neighborhoods [@Guest1999; @Dawkins2006]. It should be no suprise, then, that neighborhoods have been shown to mobilize and demobilize voters through mechanisms above-and-beyond individual characteristics [Gimpel2004; @Cho2006]. To the extent that felony disenfranchisement policies have depressive effects on turnout in the social and filial networks of the imprisoned and paroled, these effects are likely to be closely concentrated in the neighborhoods where the disenfranchised live.

## Turnout Among Formerly Disenfranchised Individuals {-}

Much of the literature discussed above has established that formerly incarcerated individuals rarely vote, even when they are no longer formally barred from doing so [@Haselswerdt2009; @Burch2011; @Meredith2015]. The causal effect of incarceration on participation is the subject of some debate within the field. Individuals who go to prison share many characterstics with lower propensity voters generally. Less educated citizens, for instance, turnout at low rates whether they have been to prison or not. In an attempt to disentangle sociodemographic characteristics from the experience of imprisonment, @Gerber2017 uses administrative data from Pennsylvania to estimate turnout rates prior to and after incarceration. They argue that the vast majority of low turnout among formerly incarcerated individuals can be explained by observable characteristics, concluding that "it appears that spending time in prison does not have large negative effects on subsequent participation" (p. 1144).

@White2019, however, indicates that interaction with the criminal justice system for individuals in the context of arrests for misdemeanor charges may have depressive effects on turnout. This finding does not necessarily conflict with @Gerber2017; as the earlier paper explains, incarceration often occurs after many other interactions with the criminal justice system. Individuals arrested for misdemeanors, on the other hand, likely reflect a much broader swath of the population --- and, therefore, individuals who may have had fewer interactions with the criminal justice system. The findings in @White2019 agree with much previous research which shows that individuals who have negative interactions with the state are less likely to participate in civic life [@Pierson1993]. @Weaver2010 argues that "contact with the institutions of criminal justice is important in structuring patterns of participation long assumed in the dominant literature to stem primarily from aspects of the individual" (p. 829).

Regardless of the precise mechanism, the low turnout among formerly incarcerated individuals is cause for concern, particularly given the racialized aspects of the criminal justice system. A criminal justice system that inflicts dire consequences on a population that has relatively little political voice is problematic. Whether or not incarceration *causes* low turnout, however, the state has a unique opportunity to craft policies that will impact individuals under formal supervision. Even if incarceration does not lead to lower turnout, policies targeting individuals caught up in the criminal justice system might still be effective at increasing turnout for this population.

Some research has been done in this area. @Meredith2015 examines the impact of ending permanent disenfranchisement in Iowa. They find that individuals who received letters explicitly informing them of their re-enfranchisement were more likely to cast ballots in the next election than those who did not. @Meredith2013, however, examines states where so-called notification laws went into effect. Although rules about eligibility did not change in these states, new policies required Departments of Corrections to notify formerly disenfranchised individuals of their re-instated voting rights. @Meredith2013 finds no effect from notification in the absense of eligibility changes.

@Gerber2014 conducted a field experiment in Connecticut in advance of the 2012 presidential election, finding that sending mailers to individuals to remind them of their voting rights was successful at increasing turnout among this population. "Whatever the participatory consequences of incarceration," they conclude, "they are not in large part impossible to overcome" (p. 924). It is not clear whether this increase in turnout is *undoing* the depressive effect of incarceration or boosting the participation of individuals whose (low) propensity to vote was unaffected by incarceration.

Executive Order 181 is expected to have increased turnout among formerly incarcerated individuals through two primary mechanisms. The first addresses the impact of negative experiences with the state. To the extent that parole officers are accurately informing their parolees of their newly restored voting rights, Executive Order 181 is likely to bring about a positive interaction between the parolee and the government. Rather than simply have one's rights restored upon completion of sentence, Executive Order 181 may lead parolees to feel explicitly invited back into the democratic process --- an invitation that may be successful at repairing some of the negative associations created through incarceration.

Secondly, Executive Order 181 is expected to dispel confusion about eligibilty and serve as a reminder to vote. As @Drucker2005 and others have detailed, many formerly incarcerated individuals are misinformed about their eligibility to cast a ballot. If confusion about eligibility leads to lower turnout, parole officers informing their parolees of their rights is likely to increase turnout. Even if confusion does not lead to lower turnout, there is some reason to believe that reminding formerly incarcerated individuals of their rights is a successful intervention for boosting turnout [@Meredith2015; @Gerber2014].

# Data {-}
## Criminal Justice Data {-}

The primary criminal justice dataset comes from a public records request filed by the author to obtain individual-level incarceration and parole records for individuals sentenced to incarceration in New York State since 1990. The data includes a host of information, including: first, middle, and last name; date of birth; class of offense; incarceration start and end dates; dates of parole; county of committment; and others. This analysis is limited to individuals incarcerated for felony offenses. Individuals convicted of misdemeanors are not disenfranchised in New York State. These data come from the New York State Department of Corrections and Community Supervision (NYSDOCCS). These data are used to determine when individuals were incarcerated or on parole, and when they finished their parole supervision.

The state does not make a unified database of parolees whose voting rights have been restored available to the public. However, the NYSDOCCS Parolee Lookup website includes a flag indicating whether someone's voting rights have been restored. By using identification numbers from the parolee data obtained from the state, I constructed a list of the individuals who were on parole and had their rights restored.^[This list was compiled by using a webscraper written in the Python language.]

## Voter File Data {-}
Most states in the United States are required to maintain files with information on all registered voters. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first, middle, and last name; date of birth; home address; vote history; and other information. The New York State Voter File also includes information on voters who were previously registered but have since been purged, either because they moved, died, or were incarcerated for a felony offense. The first section of this paper uses a snapshot of the registered voter file from April 30^th^, 2018. The second section of this paper uses a snapshot from March 3^rd^, 2019.

## Geocoding {-}
Voters' home addresses were converted to latitudes and longitudes using a geocoder provided by SmartyStreets. I then used the statistical software R to map these latitudes and longitudes to census block groups, census tracts, and city council districts using shapefiles publicly available from the Census Bureau and the City of New York.

# Effects of Felony Disenfranchisement on Neighborhood Turnout Levels {-}

## Identification of Lost Voters {-}
In this analysis, I offer a different defintion of "lost voter" than much of the literature. Many recent papers have attempted to identify relationships between the number of disenfranchised residents --- *potentially* lost voters --- and turnout. Such an approach is informative for understanding the impact of disenfranchisement. Many young men, for instance, are admitted to prison each year. If someone is incarcerated shortly after they turn 18 in an odd-numbered year, they may be incarcerated before they even have the opportunity to cast a ballot. This analytical approach is also often taken due to data constraints: most states' registered voter files provide only a snapshot of currently registered voters, making it impossible to determine voting history for individuals who are currently disenfranchised (and therefore not currently registered). Similarly, state-level data does not report the number of incarcerated individuals with a history of voting; therefore, studies that leverage variation in laws between states to estimate the effect of felony disenfranchisement rely on estimates of the total disenfranchised population, not the number of disenfranchised individuals with a history of voting.

Examining the impact of all disenfranchised voters, however, limits our ability to understand the true effects of felony disenfranchisement. As previous research has detailed, turnout rates among individuals who are incarcerated are very low even prior to incarceration. If incarcerated individuals would not have cast a ballot if they were not incarcerated, any effect of their incarceration on neighborhood turnout cannot be attributed to felony disenfranchisement *per se* but rather to their incarceration. At the local level where felony disenfranchisement rules do not vary from neighborhood to neighborhood, such a definition provides insight only into the effect of incarceration on turnout, not felony disenfranchisement.

Here, I explore whether the disenfranchisement of residents with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. Although not all lost voters would have participated if they had not been disenfranchised, past participation is an extremely strong predictor of propensity to vote [@Gerber2003]. As such, these are the individuals most likely to have been impacted not only by incarceration but specifically by felony disenfranchisement. Identifying the effects of felony disenfranchisement in neighborhoods that lost individuals with a record of voting --- individuals who would likely have cast a ballot had they been allowed --- provides insight into whether felony disenfranchisement reduces neighborhood turnout through socialization mechanisms.

To study the effect of felony disenfranchisement on voting at the local level (and its potential implications for the distribution of political power at the local level), I use turnout in the most recent non-special election for city-wide office in New York --- the mayoral election which took place on November 7^th^, 2017. Lost voters, therefore, are all individuals who were incarcerated or on parole on November 7^th^, 2017, and had cast a ballot between 2007 and 2016.

These individuals are identified by matching NYSDOCCS data to the voter file. I use first, middle, and last names, and dates of birth, to join these two datasets.^[Matching on first and last names and dates of birth can result in false positive matches, especially in large states like New York. Following @Meredith2013, I discuss the potential impact of these errors in [Appendix A].] Figure \ref{fig:citymap} shows where these lost voters lived before going to prison, with city council districts also included. There were `r comma(readRDS("./temp/lost_count_nyc.rds"))` such lost voters within New York City as of the 2017 general election, and `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` statewide.

```{r citywide-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:citymap}Lost Voters on Election Day, 2017", out.extra = '', fig.pos = "H", fig.align = "center"}
j <- readRDS("./output/city_map.RDS")
print(j)
```
The spatial concentration of lost is readily apparent. In some communities, such as Greenwich Village and Brooklyn Heights, hardly any voters were disqualified from participating in the 2017 elections. In other communities, such as Harlem and Central Brooklyn, large numbers of individuals with a demonstrated history of voting were not allowed to cast a ballot for mayor.

## Testing for Neighborhood Turnout Effects {-}
```{r table-table, , echo = F, message = F, warning = F}

```
Table \ref{tab:nhood-dist} shows the distribution of the number of lost voters by neighborhood. Most neighborhoods lost no voters at all; when a neighborhood was home to lost voters, they generally lost just one resident with a history of voting to felony disenfranchisement. More than two-thirds of block groups that lost a voter lost only one voter, while 45 percent of census tracts with a lost voter lost only one. 

```{r distribution-lost, echo = F, message = F, warning = F}
t <- readRDS("./temp/distribution_of_lost_voters.rds") %>% 
  mutate_if(is.numeric, ~ comma(ifelse(is.na(.), 0, .)))

j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:nhood-dist) Distribution of Lost Voters", linesep = "",
                  align = "c") %>% 
  row_spec(15, hline_after = T) %>% 
  kable_styling(latex_options = c("HOLD_position"))
j
```

Because the majority of neighborhoods have either zero or one lost voter, I begin testing the effect of lost voters on neighborhood turnout using a matching model. Neighborhoods (defined as census tracts and block groups) are considered treated if they were home to *at least one* lost voter in the 2017 election; they are untreated if no voters were disqualified from the election because of a felony conviction.^[It is worth noting that this definition of lost voter is highly conservative. For instance, some individuals who would have voted may have gone to prison after turning 18 but before their first general election. These "untreated" neighborhoods are likely home to individuals who would have voted if it were not for disenfranchisement policies.] I use a genetic match to match treated to untreated census tracts and block groups [@Sekhon2011], using a series of demographic and political indicators. Estimates of racial characteristics, median income, education, age, population, and share noncitizen^[The Census Bureau does not make noncitizen estimates available at the block group level. As such, block groups are assigned their census tract's share noncitizen for matching purposes.] come from the Census Bureau. Party affiliation rates come from the geocoded voter file. Registration rate is calculated by dividing the number of registered voters (calculated using the voter file) by the voting age population (estimated by the Census Bureau). I include voteshare won by the winning city council representative in 2017 as a proxy for the competitiveness of the local race.^[Where neighborhoods cross council district lines, this measure is the mean competitiveness faced by each voter in the neighborhood] Each treated block group is matched to 30 untreated block groups, and each treated census tract is matched with 10 other tracts. Matching is done with replacement. Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} present the results of these matches.

### Match Output {-}
```{r block-table-chunk, echo = F}
t <- readRDS("./temp/match_table_block_group.rds")
j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:blockgroup-table) Results of Block Group-Level Matching", linesep = "") %>% 
  add_header_above(c(" " = 1, "Means: Unmatched Data" = 2, "Means: Matched Data" = 2, "Percent Improvement" = 4), align = "c") %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
j
```

```{r tract-table-chunk, echo = F}
t <- readRDS("./temp/match_table_tract.rds")
j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:tract-table) Results of Tract-Level Matching", linesep = "") %>% 
  add_header_above(c(" " = 1, "Means: Unmatched Data" = 2, "Means: Matched Data" = 2, "Percent Improvement" = 4), align = "c") %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
j
```
At both the tract and block group level, matching results in an untreated group of neighborhoods that looks substantially like the treatment group. These tables also demonstrate the striking extent to which neighborhoods with lost voters differ from the average neighborhood in New York City. Neighborhoods with lost voters are far less white, have much lower median incomes, and a larger share of voters are registered as Democrats.

After matching neighborhoods, I use a simple regression to test whether turnout in the 2017 mayoral election was different in areas with lost voters. Census tract and block group level turnout rates are calculated using the geocoded voter file. Each voter's record indicates whether the voter participated in the 2017 general election, which are then aggregated to estimate the number of ballots cast in each neighborhood.^[The New York registered voter file does not align exactly with results reported by the city. The voter file indicates that just `r comma(readRDS("./temp/vf_ballots_17.rds"))` voters cast a ballot in the 2017 mayoral election, but the Board of Elections reports that `r comma(readRDS("./temp/mayoral_ballots.rds"))` votes were cast. In [Appendix B], I demonstrate that there is no relationship between lost voters and underreporting of cast ballots at the precinct level. These reporting anamolies are unlikely to impact this analysis.] The number of ballots cast is divided by the neighborhood's voting age population.

Much of the literature has discussed whether felony disenfranchisement is particularly demobilizing for eligible Black voters. I therefore include models which explore any potential difference in treatment effect in neighborhoods where a higher share of the population is Black. Table \ref{tab:match-reg} presents the results of these regression models. Each neighborhood is weighted by its adult population, and robust standard errors are clustered at the level of the match [@Abadie2019].
```{r tract-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/match_reg_block_group.rdata")

load("./temp/match_reg_block_group_nhb.rdata")
bg_reg <- reg_output
bg_reg_ses <- reg_output_ses

bg_reg_black <- reg_output_nhblack
bg_reg_black_ses <- reg_output_nhblack_ses

load("./temp/match_reg_tract.rdata")

load("./temp/match_reg_tract_nhb.rdata")
tract_reg <- reg_output
tract_reg_ses <- reg_output_ses

tract_reg_black <- reg_output_nhblack
tract_reg_black_ses <- reg_output_nhblack_ses

stargazer(bg_reg, bg_reg_black, tract_reg, tract_reg_black,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Tract Level"),
          header = F,
          se = list(bg_reg_ses, bg_reg_black_ses, tract_reg_ses, tract_reg_black_ses),
          notes = c("Robust standard errors (clustered at", "match level) in parentheses."),
          type = "latex",
          covariate.labels = c("D(Neighborhood Lost a Voter)", "Share Non-Hispanic Black", "D(Lost Voter) $\\times$ Share Non-Hispanic Black", "Constant"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:match-reg) Matching Regression",
          table.placement = "H",
          omit.stat = c("f", "ser"))
```
When neighborhoods are measured at the block group level, a lost voter is negatively associated with turnout in the 2017 election. In block groups with lost voters, turnout was on average `r -1*round(bg_reg[["coefficients"]][["treatTRUE"]]*100, digits = 1)` percentage points lower than in comparable block groups without lost voters. This decrease, however, appears to be entirely concentrated within Black neighborhoods. When the dummy identifying neighborhoods with lost voters is interacted with the share of the neighborhood that is Non-Hispanic Black, the basic treatment dummy becomes nonsignificant. The coefficient on the interaction between treatment and share Black indicates that treated neighborhoods that are largely Black saw turnout that was as much as `r -1*round(bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]]*100, digits = 1)` percentage points lower than similar neighborhoods without lost voters. Considering that the overall turnout rate in block groups with a lost voter was just `r round((sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap))*100, digits = 1)` percent, this effect is alarmingly high. For every 100 votes cast in a predominantly Black block group with a lost voter, as many as `r round(100 - ((100 * ((sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap)) + bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]])) / (sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap))), digits = 1)` votes went uncast. Although these variables are largely nonsignificant at the census tract level, this is not particularly surprising. If lost voters have the largest depressive effect on family members and close neighbors, it makes sense that the effect of losing a voter is locally concentrated.

### Testing Intensity Effects {-}

Matching methodologies, of course, only allow us to test the effect of being treated --- here, losing any voter for the 2017 election. The models above do not allow for different effects on turnout based on *how many* voters a neighborhood lost, potentially understating the impact of felony disenfrancisement in the most hard-hit communities. In Table \ref{tab:trad-reg} below, I adopt a standard ordinary least squares regression to investigate whether lost voters are associated with lower turnout rates in the 2017 election. This regression uses the same covariates used in the matching procedure described above. Each neighborhood is weighted by its voting age population, and robust standard errors are clustered by city council district.^[Where neighborhoods cross city council district lines, they are assigned the district in which most of their voters live for clustering purporses.]

```{r trad-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/tract_model_reg_ols.rdata")
load("./temp/bg_model_reg_ols.rdata")

stargazer(bg_model, bg_model2, tract_model, tract_model2,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Tract Level"),
          header = F,
          type = "latex",
          se = list(bg_model_ses, bg_model2_ses, tract_model_ses, tract_model2_ses),
          covariate.labels = c("Lost Voters", "Lost Voters $\\times$ Share Non-Hispanic Black", "Median Income (Thousands of Dollars)", "Percent Latino", "Percent Non-Hispanic Black", "Percent Non-Hispanic White",
                               "Percent With Some College", "Median Age", "Registration Rate", "Percent Democrats", "Percent Noncitizen", "Percent Won by City Council Representative"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:trad-reg) Standard Regression",
          table.placement = "H",
          omit.stat = c("f", "ser"),
          notes = c("Robust standard errors (clustered by city",  "council district) in parentheses."),
          omit = "boro")
```
The results presented in Table \ref{tab:trad-reg} align closely with the estimated effect from the matching model. Lost voters are generally associated with lower turnout (each missing voter in a block group reduces that neighborhood's turnout by about `r -1*round(bg_model[["coefficients"]][["lost_voters"]]*100, digits = 2)` percentage points), but Models 2 and 4 again make clear that this effect is concentrated in Black neighborhoods. In neighborhoods where most residents are Black, each lost voter is associated with a turnout decrease of up to `r -1*round(bg_model2[["coefficients"]][["lost_voters_black"]]*100, digits = 2)` percentage points. The neighborhoods most affected by felony disenfranchisement are neighborhoods where incarceration patterns overlap with Black communities.

The block groups where these depressive effects are not randomly distributed througout the city. They are highly spatially concentrated in Central Brooklyn, Eastern Queens, and Harlem. Figure \ref{fig:postest} applies the coefficient on $Lost\ Voters \times Share\ Black$ from Model 2 in Table \ref{tab:trad-reg} to the city's block groups. The estimated depressive effect is $-0.017\times Lost\ Voters \times Share\ Black$.

```{r postest-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:postest}Estimated Depressive Effect of Felony Disenfranchisement", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/dec_block_map.rds")
print(j)
```

## Discussion {-}

During the 2017 mayoral election, felony disenfranchisement laws were responsible for removing an estimated `r comma(readRDS("./temp/lost_count_nyc.rds"))` from New York City neighborhoods. The spatial concentration of these lost voters is striking: as Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} make clear, these voters were removed from neighborhoods with significantly lower incomes than the rest of the city. They were also removed from neighborhoods that were far less white than the average neighborhood. The systematic removal of voters from these neighborhoods is troubling. However, more than 900,000 votes were cast in the general election of 2017. Although felony disenfranchisement rules have enormous implications for the individuals targeted by them, it is unlikely that the removal of such a small number of voters --- concentrated though they are --- is likely to have large implications on its own.

As this analysis makes clear, however, felony disenfranchisement reaches beyond the individuals who are incarcerated. Previous literature has established that felony disenfranchisement likely impacts Black turnout at the state level. This analysis demonstrates that these demobilizing effects intersect with geographical space to systematically depress the vote in neighborhoods where voters are being sent to prison. These communities are already home to some of the poorest and most marginalized voters in the city; they are home to residents who have the least ability to influence policy through other means such as large campaign contributions. The fact that felony disenfranchisement policies appear to be disincentivizing participation on election day --- that these policies are further weakening the ability of these neighborhoods to participate in the democratic process --- is cause for alarm.

\pagebreak

# Executive Order 181 and Turnout in 2018 {-}

Prior to 2018, New Yorkers convicted of felony offenses and sentenced to prison were disenfranchised until they had completed all terms of their sentence --- their period of incarceration as well as any parole term. For New Yorkers on life parole or sentenced to life in prison, this law resulted in effective lifetime disenfranchisement. New Yorkers sentenced to felony probation, on the other hand, did not lose their voting rights.

On April 18^th^, 2018, Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. Such a move was of course good for the communities in which parolees live; as discussed above, the disenfranchisement of voters has large spillover effects in the neighborhoods in which these lost voters are concentrated. The change in policy is also beneficial for felony probationers: despite the fact that probationers do not formally lose their voting rights, there is evidence that confusion around the law contributes to *de facto* disenfranchisement among probationers [@Drucker2005].^[I've also talked with folks on probation whose probation officers told them they were disenfranchised. Not sure if it's appropriate to mention anecdotes like these, or to just cite to the literature] The Executive Order is a promising step: by changing the policy to allow all New York citizens living in their communities to cast a ballot, the move has the potential to both re-enfranchise the nearly 30,000 New Yorkers on parole living in the community and to clarify the rules about who is eligible to vote.

Of course, re-extending the right to vote is only the first step. As previous literature has established, interactions with the criminal justice system leaves residents less likely to vote in the future [@White2019]. As Traci Burch [-@Burch2011] and others have shown, moreover, turnout rates among the formerly incarcerated are extremely low. Formal disenfranchisement policy, the literature has made clear, is just one piece of an interlocking system that serves to disenfranchise minority and marginalized voters. To address *only* the formal laws contributing to disenfranchisement without also interrogating the efficacy of any policy change risks leaving much of the system of effective disenfranchisement undisturbed.

There is reason to believe that the policy change may increase the political participation of formerly disenfranchised individuals. Prior to the policy change, formerly incarcerated individuals had their voting rights restored automatically upon the completion of their parole term. Under the Executive Order, however, parole officers were required to explicitly inform their supervisees of their newly restored voting rights. It is possible that having a representative of the government tell parolees of their rights was an effective encouragement for these newly re-eligible voters to cast a ballot.

In the analysis below, I examine the effect of Executive Order 181 on individuals who finished parole before October 10^th^, 2018. These are the individuals who would have been eligible to cast a ballot whether or not the Executive Order was signed. This is, of course, only a subset of the individuals who were impacted by Executive Order 181 --- many of the individuals who had their voting rights restored were still on parole on Election Day, and therefore were eligible to vote only because of the rules change. However, by focusing on individuals who would have been eligible to vote either way, we can test whether rights restoration prior to parole discharge serves as an effective encouragement to cast a ballot for this population.

## Identifying Parolees Whose Rights Were Restored {-}
Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored. By using the identification number provided from the parolee public records request, I was able to identify individuals who had their voting rights restored.^[Not all parolees listed in the public records request data are included in the lookup tool. For individuals who finished parole between January 1^st^, 2018, and April 17^th^, 2018, 1.0 percent are not in the lookup tool. For those discharged from parole between April 18^th^, 2018, and January 13^th^, 2019 (the latest date of the parole records), 1.2 percent of individuals are not found in the lookup tool.]

## Trends in Turnout {-}
Before analyzing turnout in the 2018 midterms, I begin by examining turnout in the 2016 election. It is possible that individuals discharged from parole shortly before a federal election are more likely to cast a ballot than individuals discharged earlier, whether or not their voting rights were restored. However, as Figure \ref{fig:to-16} makes clear, individuals discharged from parole in the final months before the 2016 presidential election were not substantially more likely to cast a ballot in the election than individuals discharged earlier. The longer an individual has been off of parole, the more likely he is to cast a ballot. For instance, of the individuals last discharged from parole in 2010, `r percent(readRDS("./temp/turnout_in_16_finished_10.rds"), accuracy = 0.1)` cast a ballot in the 2016 election, while just `r percent(readRDS("./temp/turnout_in_16_finished_15.rds"), accuracy = 0.1)` of those last discharged from parole in 2015 did so.^[Figure \ref{fig:to-16} plots individuals' turnout by the last date of discharge from parole. Therefore, individuals discharged from parole in 2010 who reoffended and were discharged from parole again in 2015 are included only in 2015.] Figure \ref{fig:to-16} plots turnout rates by month of parole discharge. A quadratic curve is fitted (weighted by the number of individuals discharged each month), along with a 95 percent confidence band. This curve is fit on monthly data running from January, 2010 through April, 2016, and extended through October, 2016.

```{r to-16-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-16}Turnout in 2016 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to16_chart.rds")
print(j)
```
Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January, 2012 through April, 2018, and extended through October, 2018.
```{r to-18-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart.rds")
print(j)
```
Figure \ref{fig:to-16} does not indicate that individuals who were discharged from parole shortly before the 2016 presidential election were more likely to cast a ballot than individuals discharged earlier in the year. Figure \ref{fig:to-18}, on the other hand, indicates that New Yorkers discharged from parole in the months leading up to the 2018 election --- many of whom had their rights restored while they were still on parole --- were more likely to participate than those discharged earlier in the year. However, Figures \ref{fig:to-16} and \ref{fig:to-18} are noisy and do not prove that Executive Order increased turnout.

## Individual-Level Turnout Regressions {-}

When considered over a multi-year period, the enactment of Executive Order 181 cannot be understood as a natural experiment. The longer an individual has been off of parole, the more likely she is to cast a ballot, but only individuals recently discharged from parole were eligible to have their voting rights restored prior to discharge. For a true natural experiment to hold, an individual's probability of being "assigned" to treatment (here, discharged from parole after the Executive Order went into effect) must be uncorrelated with the outcome of interest (propensity to vote). Figures \ref{fig:to-16} and \ref{fig:to-18} indicate that this is not the case when considering individuals discharged from parole over multiple years.

However, the relationship between time-off-parole and propensity to vote is far weaker in the short term. Figure \ref{fig:to-18-recent} indicates that parole discharge date and turnout rates in the 2018 midterm election are not correlated when we limit the analysis to individuals discharged in 2017 or 2018.

```{r to-18-recent-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18-recent}Turnout in 2018 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart_recent.rds")
print(j)
```
Formalizing this chart into an individual-level logit model demonstrates that time since discharge is not correlated with turnout in the short run. The models in Table \ref{tab:to-18-logit-short} include individuals last discharged from parole between January 1^st^, 2017, and April 17^th^, 2018.

```{r logit-reg-chunk-short, echo = F, results = "asis", message = F}
load("./temp/short_term_to.rdata")

stargazer(date_to_model_notime, date_to_model_time,
          header = F,
          type = "latex",
          covariate.labels = c("Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:to-18-logit-short) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "X", "X"),
                         c("Felony Class Fixed Effects", "X", "X")),
          table.layout = "-dm#-t-a-s-n")
```

The inclusion of time controls in Table \ref{tab:to-18-logit-short} increases the AIC. A Chi-squared test confirms that the model is not improved when controls for time are included.^[[Appendix A] provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.] When we look only at individuals recently discharged from parole, the length of time an individual has been off parole is not associated with his propensity to vote.

Although Governor Cuomo signed the Executive Order on April 18, 2018, an examination of the individuals whose rights were ultimately restored indicates that the program did not go into full effect until later in May. Figure \ref{fig:restoration-chart} plots the share of individuals discharged from parole each day in the spring of 2018. It is not until May 21^st^ that the majority of discharged parolees had their voting rights restored prior to their discharge.
```{r restoration-chart-chunk, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:restoration-chart}Share of Discharged Parolees Whose Voting Rights Were Restored Prior to Discharge", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/restoration_plot.rds")
print(j)
```

Although it perhaps seems counterintuitive to throw out many of our observations, ignoring all individuals last discharged from parole prior to 2017 give us an important asset: it allows us to conceptualize Executive Order 181 as a natural experiment. Whether an individual was discharged from parole before or after May 21^st^, 2018, is akin to a randomly assigned treatment. Any observed difference in turnout between individuals discharged before and after late May can therefore be attributed to the Executive Order.

In Table \ref{tab:to-18-logit}, I present the results of an individual-level logistic regression exploring whether individuals who were discharged on or after May 21^st^, 2018, turned out at higher rates than those discharged earlier. The models include all individuals discharged from parole between January 1^st^, 2017, through October 10^th^, 2018 (the registration deadline in New York State).

```{r logit-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/individual_turnout_18_itt.rdata")

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Discharged After EO 181)", "D(Male)", "Age (Years)",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:to-18-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```

Model 1 in Table \ref{tab:to-18-logit} formalizes the trend presented in Figure \ref{fig:to-18-recent-chart} by controlling only for whether an individual was discharged after Executive Order 181 went into effect. Model 2 also controls for individual-level characteristics: sex, age on November 6^th^, 2018, county of incarceration, and race. Model 3 adds sentence-specific information to Model 2: the number of counts in the individual's most recent sentence, the amount of time they spent on parole, and the class(es) of felony for which they were convicted. Table \ref{tab:to-18-logit} makes clear that formerly incarcerated men were far less likely to vote than formerly incarcerated women; that older formerly incarcerated individuals were more likely to cast a ballot; and individuals who spent longer on parole were more likely to participate in the midterm election.

Each model also indicates that individuals discharged from parole after the Executive Order went into effect were more likely to cast a ballot than those discharged earlier. Exponentiating the coefficients on D(Discharged After EO 181) indicates that Executive Order 181 raised the turnout rate among formerly disenfranchised individuals by between `r round((exp(model1[["coefficients"]][["finished_postTRUE"]]) - 1) * 100, digits = 1)` and `r round((exp(model3[["coefficients"]][["finished_postTRUE"]]) - 1) * 100, digits = 1)` percent.

## Instrumental Variables Approach {-}

Table \ref{tab:to-18-logit} indicates that Executive Order was successful at increasing turnout among all formerly disenfranchised individuals. Although this is important information for policymakers and advocates hoping to increase the political representation of formerly incarcerated individuals as a whole, it does not shed light on the extent to which having one's voting rights restored prior to discharge from parole impacts one's propensity to vote. To answer that question, we must specifically control for whether an individual did have his rights restored before he was discharged from parole.

An individual's propensity to vote can be expressed using the following equation:
$$ Y_i =  b_0 + b_{1}X_{1i} + b_{2}X_{2i} + \epsilon_{i} $$
Where *Y~i~* is 1 if individual *i* cast a ballot, *X~1i~* is the probability that individual *i* was eligible to have his rights restored prior to discharge from parole, and *X~2i~* is 1 if individual *i* actually had his rights restored prior to discharge from parole. Given that we cannot observe *X~1i~*, we could simply choose to ignore it. Doing so, however, will result in consistent regression coefficients only if *X~1i~* and *X~2i~* are uncorrelated or if *b~1~* is uncorrelated with an individual's propensity to vote. Neither of these assumptions are valid: an individual's eligibility to have her rights restored is almost certainly correlated with whether or not they actually were restored. Moreover, eligiblity for rights restoration is likely correlated with propensity to vote. Non-citizens, for instance, were not eligible for rights restoration under EO181 because non-citizens cannot vote in New York State. Similarly, individuals who were re-arrested while on parole were not eligible to have their rights restored. Citizenship status and recency of arrest are both highly correlated with propensity to vote.

Our variable of interest, therefore, is likely correlated with our regression error. Such a correlation inhibits our ability to simply regress turnout on a dummy indicating whether an individual's rights were restored. The standard way of dealing with such a problem is an instrumental variables approach. Such an approach has been widely used in the context of voter turnout [e.g. @Gerber2000; @Green2012]. A valid instrumental variable must satisfy two criteria: it must be correlated with the right-hand-side (endogenous) variable of interest, and it must be uncorrelated with the error of the regression.

In this case, such a variable is readily at hand. The likelihood that an individual had his voting rights restored is a function of whether he was discharged from parole after Executive Order 181 took effect. As demonstrated above, date of discharge from parole is not correlated with propensity to vote when we limit our analysis to individuals discharged from parole in 2017 or 2018. A dummy variable indicating whether an individual was discharged after the Executive Order, therefore, satisfies the criteria for an instrumental variable.

Table \ref{tab:iv-models} presents the two-stage least-squares regression estimates exploring the turnout impact of rights restoration. As discussed above, the instrumental variable used in these regressions measure whether an individual was discharged from parole after Executive Order 181 went into effect. These models are estimated using an OLS sepcification. Although the dependent variable remains dichotomous (as do both the independent variable of interest and the instrument), @Angrist2008 recommends using OLS in instrumental variables analyses even under such conditions.

```{r iv-reg-chunk-short, echo = F, results = "asis", message = F}
load("./temp/iv_individual_turnout_18.rdata")

summ.fit1 <- summary(iv1, diagnostics = T)
summ.fit2 <- summary(iv2, diagnostics = T)
summ.fit3 <- summary(iv3, diagnostics = T)

stargazer(iv1, iv2, iv3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Rights Restored)", 
                               "D(Male)", "Age (Years)",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:iv-models) Second-Stage Regression",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X"),
                         c("\\hline \\\\[-1.8ex]"),
                         c(rownames(summ.fit1$diagnostics)[1],
                           round(summ.fit1$diagnostics[1, "p-value"], 2),
                           round(summ.fit2$diagnostics[1, "p-value"], 2),
                           round(summ.fit3$diagnostics[1, "p-value"], 2)),
                         c(rownames(summ.fit1$diagnostics)[2],
                           round(summ.fit1$diagnostics[2, "p-value"], 2),
                           round(summ.fit2$diagnostics[2, "p-value"], 2),
                           round(summ.fit3$diagnostics[2, "p-value"], 2))),
          omit.stat = "ser",
          table.layout = "-dm#-t-as-n",
          notes = c("Table reports p-values for Weak Instruments",
                    "and Wu-Hausman tests.",
                    "First-stage equations use dummy indicating",
                    "whether individual was discharged after the",
                    "Exectutive Order went into effect. The",
                    "first-stage equations also include",
                    "covariates in models 2 and 3."))
```

Table \ref{tab:iv-models} indicates that using discharge date of on or after May 21^st^ as an instrument for rights restoration is warranted. The p-values of the Wu-Hausman tests hover just below 0.05, while the Weak Instruments test is highly significant in each model.

Model 3 in Table \ref{tab:iv-models} indicates that having one's rights restored caused an individual's propensity to vote to increase by `r round(summary(iv3)[["coefficients"]][2,1] * 100, digits = 2)` percentage points. Approximately `r round(readRDS("./temp/to_restored.rds") * 100, digits = 2)` percent of individuals who had their rights restored cast a ballot in 2018, indicating that Executive Order 181 increased the propensity to vote by around `r round(summary(iv3)[["coefficients"]][2,1] / (readRDS("./temp/to_restored.rds") - summary(iv3)[["coefficients"]][2,1]) * 100, digits = 1)` percent. Unsurprisingly, this estimate is higher than the estimates reported in Table \ref{tab:to-18-logit}. Because Executive Order 181 theoretically only impacts individuals whose rights were actually restored prior to discharge, the magnitude of the estimated effect on this subgroup must be larger than the group taken as a whole.

## Discussion {-}
As discussed above, the effect of incarceration on individuals' propensity to vote is an open question for the literature. What is known, however, is that individuals who have been to prison were likely to vote at very low rates prior to incarceration, and continue to vote at very low rates after incarceration. Whether incarceration reduces political participation or not, increasing the propensity to vote of individuals who have been to prison is a laudable goal. Prisoners overwhelmingly come from marginalized communities with much to gain from policy. Executive Order 181 appears to have substantially boosted turnout among formerly incarcerated individuals through two mechanisms. The first is obvious: parolees who would not have been eligible to vote absent the Executive Order were allowed to cast ballots in the midterm elections. The second mechanism is less obvious. Individuals who finished parole between May 21^st^, 2018, and  October 10^th^, 2018, would have been eligible to participate in the election even if the governor had not signed the Executive Order. The EO 181, however, appears to have boosted turnout even among individuals whose eligibility to vote was not directly impacted by the rules change. The increase in turnout was substantial: Executive Order 181 increased turnout among individuals who had their rights restored by more than 40 percent.

The precise mechanism through which the rules change increased turnout among these individuals is not clear. It is possibly a social mechanism: having an officer of the state affirm one's eligibility (and therefore reunion with the body politic) could be responsible for the increase. It could be due to better information: individuals whose rights were restored were likely far more confident of their eligibility to vote than other formerly incarcerated individuals. It could also be an issue of timing: individuals whose rights were restored in May of 2018 but were not discharged from parole until early October had four additional months to register to vote than they would have absent the Executive Order. In reality, the turnout boost is likely due to a combination of different factors.

Despite the substantial increase in turnout thanks to Executive Order 181, turnout among formerly incarcerated individuals remained stubbornly low in the 2018 midterm elections. Just 3.2 percent of individuals who had their rights formally restored and finished parole before the registration deadline. Although the successes of Executive Order 181 should be celebrated, more must be done to encourage formerly incarcerated individuals to participate in the political process.

\newpage
# References {-}
<div id="refs"></div>

\newpage
# Appendix A {-}

Matching across administrative data sets is not a perfect science. Despite having high-quality low-level data from both New York State Department of Corrections and Community Supervision (DOCCS) and the New York State Board of Elections, I cannot be sure that I am correctly identifying formerly incarcerated, registered voters. The results of the analyses in Table \ref{tab:match-reg} give some reason to believe that there are few false negatives. If there were many false negatives (lost voters who are not identified in this analysis), "treated" neighborhoods would be incorrectly categorized as "untreated" neighborhoods. If this were a widespread problem, we would expect to see little difference between treated and control neighborhoods. To the extent that lost voters go unidentified by this analysis, it is likely that the results presented in Tables \ref{tab:match-reg} and \ref{tab:trad-reg} are conservative by including treated neighborhoods in the control group. Directly testing for false negatives, however, is difficult. If a voter changes her name before being incarcerated, for instance, our analysis will fail to capture her.

Testing for the presence of false positives is slightly easier. One way to do so is by constructing false records [@Meredith2013]. By slightly changing the birthdates in one set of administrative data, I can estimate how frequently DOCCS records inaccurately match with voter records. To do so, I shift the dates of birth in the DOCCS data by 35 days. I then re-merge the DOCCS records (with "false" birth dates) against the registered voter file.

Table \ref{tab:change-dobs} shows the results of these permutations. The first row shows the matches with the raw data. The second row shows the results of adding 35 days to the date of birth in each DOCCS record, and the final row shows the result of subtracting 35 days from these records.

```{r shift-dobs-chunk, echo = F}
less_35 <- scales::comma(nrow(readRDS("./temp/matched_ids_in_17_m35.rds")))
true <- scales::comma(nrow(readRDS("./temp/matched_ids_in_17.rds")))
plus_35 <- scales::comma(nrow(readRDS("./temp/matched_ids_in_17_p35.rds")))


table <- data.frame(group = c("Actual Birthdate", "Birthdate + 35 Days", "Birthdate - 35 Days"), count = c(true, plus_35, less_35))

j <- table %>% 
  mutate_all(linebreak) %>% 
  knitr::kable(booktabs = T, caption = "(\\#tab:change-dobs) Results of Shifting Birthdates", linesep = "", escape = F,
               col.names = linebreak(c("Group", "Number of Matches Between\nDOCCS and Voter File Records")),
               align = "c") %>% 
  kable_styling(latex_options = c("HOLD_position"),
                position = "center")
j

```
Table \ref{tab:change-dobs} indicates that there are likely some false positives in our matches. It is not surprising that in a large state like New York, full names and dates of birth fail to uniquely identify individuals. However, the rate of false positives is quite low: this analysis indicates that there is a `r percent(as.integer(max(less_35, plus_35)) / as.integer(gsub(",", "", true)), accuracy = 0.01)` rate of false positives. It is unlikely that such a small incidence of false positives meaningfully impacts this analysis.

\newpage
# Appendix B {-}

As noted above, the turnout figures reported in the registered voter file do not align with the results reported by the New York City Board of Elections. According to the official results, `r comma(readRDS("./temp/mayoral_ballots.rds"))` ballots were cast in the 2017 mayoral election. The registered voter file reports that just `r comma(readRDS("./temp/vf_ballots_17.rds"))` voters cast a ballot. This is not necessarily evidence of poor management by the Board of Elections. Some voters choose not to make their voter registration information publicly available (such as domestic violence survivors and officers of the court).^[This is not to say that there is *not* evidence of mismanagement by the BOE. For instance, the author is incorrectly marked as not participating in the 2017 mayoral primary.] Where more ballots are recorded by the BOE than in the voter file, turnout calculated from the voter file will be artificially lower. If the voter file undercount rate is systematically worse in neighborhoods with lost voters, this would pose a serious challenge to the validity of the results reported in the body of this paper.

To test whether there is a relationship between lost voters and voter file undercount, I examine the undercount rate at the precinct level. The voter file indicates the home precinct of each voter, and the Board of Elections publishes election results at the precinct level. Precincts are assigned the demographics of the block group in which they are situated. Where precincts cross block group lines, the precinct is assigned the characteristics of all block groups in which it is located weighted by the number of voters from each block group. The undercount rate is calculated as the number of votes derived from the voter file divided by the number reported in BOE results. Robust standard errors are clustered at the assembly district level.

```{r undercount-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/compare_ballots_off_results.RData")

stargazer(ed_reg,
          header = F,
          type = "latex",
          se = list(ed_comp_ses),
          covariate.labels = c("Lost Voters", "Median Income (Thousands of Dollars)", "Percent Latino", "Percent Non-Hispanic Black", "Percent Non-Hispanic White",
                               "Percent With Some College", "Median Age", "Registration Rate", "Percent Democrats", "Percent Noncitizen", "Percent Won by City Council Representative"),
          dep.var.labels = c("Undercount Rate"),
          title = "(\\#tab:uc-reg) Registered Voter File Ballot Undercount",
          table.placement = "H",
          omit.stat = c("f", "ser"),
          notes = c("Robust standard errors (clustered by", "assembly district) in parentheses."),
          omit = "boro")
```

As Table \ref{tab:uc-reg} makes clear, there is very little relationship (p = 0.47) between the number of lost voters in a precinct and the ballot undercount rate. Although researchers should be somewhat wary using turnout rates derived from New York State's registered voter file, there is no evidence that the reporting error impact this analysis.

\newpage
# Appendix C {-}

In the [Instrumenting Rights Restoration] section of this paper, I argue that being discharged from parole in the final months leading up to an election is uncorrelated with propensity to vote. Table \ref{tab:to-16-logit} demonstrates that individuals discharged between May 21^st^ -- October 14^th^, 2016, did not participate at different rates in the 2016 presidential election than other formerly incarcerated individuals. Table \ref{tab:to-16-logit} includes all individuals last discharged from parole between January 1^st^, 2015, and October 14^th^, 2016.
```{r logit-reg-chunk-16, echo = F, results = "asis", message = F}
load("./temp/individual_turnout_16.rdata")

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Discharged on or after May 21\\textsuperscript{st}, 2016)", "Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2016 Election"),
          title = "(\\#tab:to-16-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```
Although turnout was generally higher in 2016 than in 2018 (reflecting statewide higher turnout thanks to the presidential contest), there is no evidence that being discharged in the summer of 2016 was associated with an individual's propensity to cast a ballot (the p-value on the coefficient of interest exceeds 0.5 in each model). The nonsignificant results from 2016 provide strong corroboration that discharge date serves as an effective instrument for rights restoration.